What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-14 android-stacks.tex
Several smaller textual improvements to introduction
And removed paragraph about private issues,
as we don't look into those.

Often exception handling constructions may lead the developers into believing that
by just re-throwing the exceptions they can forget about the exceptional
situations during the development of the "happy path". This "ignore-for-now"
approach may turn the exception handling into a generalized "goto"
mechanism~\cite{mandrioli1992advances} making the program more complex and even
less reliable. This behaviour may lead to the well known uncaught exceptions
problem~\cite{jo2004uncaught}, one of the main causes of application crashes.

Moreover, some languages use exceptions to represent programming mistakes
(out-of-bounds array index, division-by-zero, access to null pointers). In these cases,
exceptions are implicitly signaled by the runtime environment when such
conditions occur. One of such implicit exceptions (caused by data conversion
from a 64-bit floating point to a 16-bit signed integer) was the responsible for
the famous failure on Ariane 5's first test flight~\cite{lions1996ariane} ---
leading to the rocket self-destructing 37 seconds after launch and a loss of 500
million dollars. Besides Ariane 5, several applications crash everyday due to
uncaught exceptions~\cite{jo2004uncaught}. Not using exceptions carefully
may have the opposite effect of the initial intention of improving
system robustness.

In Java, when the program fails due to an uncaught exception, it automatically
terminates, while the system prints a stack trace to the console. A stack trace
includes a detail message and the execution stack frame when the exception
occurred. A typical Java stack trace consists of an ordered list of methods that
were active on the call stack before the exception has occurred. Stack traces
are a useful source of information about system crashes, and is often used to
support developers in debugging~\cite{schroter2010stack}. Moreover, information
from stack traces combined with crash and bug reports can be used to support bug
classification and clustering~\cite{wang2013improving, kim2011crash,
dhaliwal2011classifying}, fault-proneness prediction
models~\cite{kim2013predicting} and even automated bug
fixing~\cite{sinha2009fault} tools.

However, all these works focus on the analysis of a single system at a time. In
our current context, on which we have plenty access to the information available
on open-source repositories like GitHub initial questions arise: What is the
prevalence of exception stack traces on reported issues? What are the common
characteristics of such stack traces?

@@

To answer these questions, we conduct an exploratory study of stack traces in
the wild. Using a custom tool called ExceptionMiner, that was specifically
developed to for this study, we mined stack traces from issues across all Java
projects on GitHub. Overall, we analyzed 356,057 issues from 16,836 projects,
from which 28,800 stack traces were extracted and used as a source of
information for our analysis. The stack trace analysis  was augmented by
additional information extracted using byte code and source code analysis done
on a carefully selected subset of our initial sample (482 Android projects).
%This study was divided in two phases: in the fist phase all the issues of all
%Java projects available on GitHub were analyzed and the embedded stack traces
%extracted using ExceptionMiner.

@@

This study is built on the following assumption: the stack traces reported on
issues contain relevant information related to system crashes. Several studies
and techniques have also based on this assumption
(e.g.,~\cite{sinha2009fault}~\cite{dhaliwal2011classifying}~\cite{kim2013predicting})
and we build on it as well. Some outcomes were consistently detected through
this large scale analysis of exception stack traces in the wild, such as:

@@

  \item  A multitude of programming mistakes as the main causes of stack traces.

@@

  \item  Odd exception chains (e.g., Error wrapping checked and runtime
    exceptions and vice-versa). As an indication that the purpose of Java hybrid
    exception model may not have been adequately used.

@@

  \item  It introduces ExceptionMiner, a tool developed to support the
    analysis.

@@

The contributions of this work allow for developers of Java applications
to familiarize with the most common reported exceptions mis-uses which is the
first step to help developers to avoid making them; (ii) designers of languages
to consider ways of reducing the abundance of NullPointerExceptions, (iii) and
tool designers to consider developing tools that enable a cross-project defect
analysis. We strongly believe that, specially in the open-source environment,
faults are not to be hidden in a private bug issue. Faults should be shared and
discussed, so that a developer can learn from other projects mistakes.
Currently, the search facilities of repositories are very limited, and the
ExceptionMiner tool is a contribution in this direction.

%GGG: Personally, I hate this paragraph in all papers as it usually reads like a space-filler
%RSC: If needed we can remove it.
The remainder of this paper is organized as follows. Section 2 presents a
background on exception handling mechanism. Section 3 presents the study design.
Section 4 reports study findings. Section 5 provides further discussions and insights.
Section 6 presents the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 presents our conclusions and
directions for future work.

>>>>

Often exception handling constructions may lead the developers into believing that
by just re-throwing and possibly logging the exceptions they can forget about the exceptional
situations during the development of the "happy path". This "ignore-for-now"
approach may turn the exception handling into a generalized "goto"
mechanism~\cite{mandrioli1992advances} making the program more complex and even
less reliable. This behaviour may lead to the problem of uncaught exceptions~\cite{jo2004uncaught}, one of the main causes of application crashes.

Moreover, some languages use exceptions to represent programming mistakes
(array index out-of-bounds, division-by-zero, dereferencing null pointers). In these cases,
exceptions are implicitly signaled by the runtime environment when such
conditions occur. One of such implicit exceptions (caused by data conversion
from a 64-bit floating point to a 16-bit signed integer) was responsible for
the infamous failure on Ariane 5's first test flight~\cite{lions1996ariane} ---
leading to the rocket self-destructing 37 seconds after launch and a loss of 500
million dollars. Besides Ariane 5, several applications crash everyday due to
uncaught exceptions~\cite{jo2004uncaught}. Not using exceptions carefully
may have the opposite effect of the initial intention of improving
system robustness.

In Java, when the program fails due to an uncaught exception, it automatically
terminates, while the system prints a stack trace to the console. A stack trace
includes an explanatory message and the execution stack frame when the exception
occurred. A typical Java stack trace consists of an ordered list of methods that
were active on the call stack before the exception occurred. Stack traces
are a useful source of information about system crashes, and are often used to
support developers in debugging~\cite{schroter2010stack}. Moreover, information
from stack traces combined with crash and bug reports can be used to support bug
classification and clustering~\cite{wang2013improving, kim2011crash,
dhaliwal2011classifying}, fault-proneness prediction
models~\cite{kim2013predicting} and even automated bug
fixing~\cite{sinha2009fault} tools.

However, all existing research has focused on the analysis of a single system at a time. In
our current context, on which we have access to wealth of information available
on open-source repositories like GitHub additional questions arise: What is the
prevalence of exception stack traces on reported issues? What are the common
characteristics of such stack traces?

@@

To answer these questions, we conduct an exploratory study of stack traces in
the wild. Using a custom tool called ExceptionMiner, which was specifically
developed for this study, we mined stack traces from issues across all Java
projects on GitHub. Overall, we analyzed 356,057 issues from 16,836 projects,
from which 28,800 stack traces were extracted and used as a source of
information for our analysis. The stack trace analysis  was augmented by
additional information extracted using byte code and source code analysis done
on a carefully selected subset of our initial sample (482 Android projects).
%This study was divided in two phases: in the fist phase all the issues of all
%Java projects available on GitHub were analyzed and the embedded stack traces
%extracted using ExceptionMiner.

@@

This study builds upon the following assumption: the stack traces reported on
issues contain relevant information related to system crashes. Several studies
and techniques have also taken this assumption as starting point
(e.g.,~\cite{sinha2009fault}~\cite{dhaliwal2011classifying}~\cite{kim2013predicting}). Some outcomes were consistently detected through
our large scale analysis of exception stack traces in the wild, such as:

@@

  \item  A multitude of programming mistakes which are the main causes of stack traces.

@@

  \item  Odd exception chains (e.g., Error wrapping checked and runtime
    exceptions and vice-versa), indicating that the purpose of Java's hybrid
    exception model may not have been adequately used.

@@

  \item  It introduces ExceptionMiner, a tool developed to support cross-repository stack trace
    analysis.

@@

These contributions are relevant to:
(i) developers of Java applications
to familiarize with the most common reported exceptions mis-uses which is the
first step to help developers to avoid making them; 
(ii) designers of languages
to consider ways of reducing the abundance of NullPointerExceptions;
and (iii)
tool designers to consider developing tools that enable a cross-project defect
analysis.

% AVD: We have not looked at private issues, and the tool is described above.
%We strongly believe that, specially in the open-source environment,
%faults are not to be hidden in a private bug issue. Faults should be shared and
%discussed, so that a developer can learn from other projects mistakes.
%Currently, the search facilities of repositories are very limited, and the
%ExceptionMiner tool is a contribution in this direction.

%GGG: Personally, I hate this paragraph in all papers as it usually reads like a space-filler
%RSC: If needed we can remove it.
This paper is organized as follows. Section 2 presents a
background on exception handling mechanism. Section 3 presents the study design.
Section 4 reports study findings. Section 5 provides further discussions and insights.
Section 6 presents the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 presents our conclusions and
directions for future work.
